---
title: "Exercise Sheet 1 -- Data Mining <BR> Wirtschaftsinformatik, HTW Berlin"
author: "Martin Spott"
date: "last revision `r format(Sys.Date(), format='%d.%m.%Y')`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
install.packages("ggplot2")
install.packages('knitr', dependencies = TRUE)
library(ggplot2)
library(knitr)
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```


The exercises are about linear regression. We use the data set `01_heights_weights_genders.csv` which you can download from Moodle. It contains the weight (in american pounds) and height (in inches) of 5000 women and 5000 men.


The exercises will help you revise basic concepts of R, do descriptive statistics with visualisations and build linear regression models. Download the data set `01_heights_weights_genders.csv` from Moodle. It contains the weight (in american pounds) and height (in inches) of people. 
R comes with many functions for data visualisation like `plot`, `hist` and `boxplot`, however I recommend to have a look at other graphing libraries that are more powerful, flexible and produce better looking graphs, in particular *ggplot2*. Install it once with `install.packages("ggplot2")` and load it with `library(ggplot2)` whenever you start a new R session. 

A good starting point for using *ggplot2* is the [Cookbook for R](http://www.cookbook-r.com), specifically Section 8 on [Graphs](http://www.cookbook-r.com/Graphs/). A link to the more comprehensive book *R Graphics Cookbook* by Winston Chang can be found on the same page. 

Other useful graphing packages include *lattice* and *plotly*. 


## Exercise 1.1

a) Import the data using the function `read.csv()` and assign it to a variable called `weight_df` (a data frame) using the following code fragment. You may have to specify the path of the file. 
    ```{r, eval=FALSE}
    weight_df  <- read.csv("./01_heights_weights_genders.csv")
    ```
b) Explore the data frame `weight_df` using functions like `str()`, `dim()`, `names()`, `head()` and `View()`. How many columns has the data, what are the column names, how many rows, what are the data types etc?
c) Scale the columns for *height* and *weight* to use metric measures *cm* and *kg*: use *1 inch = 2.54 cm* and *1 kg = 2.2 pound*. Hint: A column in a data frame can be addressed using e.g. `weight_df$Height`. Please remember that R distinguishes upper and lower case. 
d) Explore the value ranges of the scaled columns *height* and *weight* using `summary()`, box plots and histograms. Distinguish between men and women. Hint: Subsets can be produced using `subset(weight_df, Gender == "Male")` or `weight_df[weight_df$Gender == "Male",]`. Alternatively, `Gender` can be used to define the colour in a plot.  
e) Produce a scatter plot with *height* on the x-axis and *weight* on the y-axis. Add descriptive labels to all the axes and give it a title. Again distinguish between men and women using separate plots or use colour. 


## Exercise 1.2

a) Find out how to build a linear regression model using the function `lm()` (also see the example in b).  
b) Build linear models of the data using *height* as input (independent variable) and *weight* as output (dependent variable) for men, women and all. Assign the result to the variables `weight_lm_m`, `weight_lm_f`, `weight_lm_all`. Example:
c) Explore the data structures `weight_lm_x`, e.g. use `names(weight_lm_x)` to learn about the columns (with `x` being one of `m`, `f`, `all`). Refer to the help pages of `lm` to find out what they mean.  
d) Add the regression lines to the scatter plots from Exercise 1 e). If you used the basic `plot` function in Exercise 1.1 e) give `abline(weight_lm_x, col="red")` a try and look at the help pages for more information. `ggplot2` offers quite convenient functions for linear regression lines as well.  
e) Compare the three regression lines and interpret the differences.  

## Exercise 1.3

a) Use `summary(weight_lm_x)` to explore the three linear models:
    * their residuals 
    * the coefficients of the model with standard error, t- and p-values of the statistical hypothesis test
    * residual standard error and R-squared (squared correlation)
    Interpret the p-values in terms of rejecting or not rejecting the null hypothesis of a parameter being zero.  
b) Look at the confidence intervals of the linear models using `confint()`. Change the confidence levels from the default 0.95, see how the confidence intervals change and explain why (use the parameter `level = ...`).



## Solutions

### Exercise 1.1a

```{r}
weight_df  <- read.csv("01_heights_weights_genders.csv")
```



### Exercise 1.1b

```{r}
head(weight_df)
names(weight_df)
dim(weight_df)
str(weight_df)
# columns = 3 (gender, height, weight)
# rows = 1000
# data types = (height, num; gender, categorical; weight, num)
```

### Exercise 1.1c

```{r}
weight_df$Height <- weight_df$Height * 2.54
# TASK: transform weight into kg accordingly
weight_df$Weight <- weight_df$Weight * 2.2
```

### Exercise 1.1d

```{r}
weight_df_m <- weight_df[weight_df$Gender == "Male",]

# TASK: do the same for women's data 
weight_df_f <- weight_df[weight_df$Gender == "Female",]

boxplot(weight_df_f$Height, weight_df_m$Height, names = c("women", "men"), main = "height")
hist(weight_df_f$Height, main = "histogram of women's height", xlab = "women's height")
hist(weight_df_m$Height, main = "histogram of men's height", xlab = "men's height")

# TASK: do the same for Weight
boxplot(weight_df_f$Weight, weight_df_m$Weight, names = c("women", "men"), main = "weight")
hist(weight_df_f$Weight, main = "histogram of women's weight", xlab = "women's weight")
hist(weight_df_m$Weight, main = "histogram of men's weight", xlab = "men's weight")

# These are alternative versions using ggplot2
library(ggplot2)
ggplot(weight_df, aes(x=Gender, y=Weight, fill=Gender)) + 
  geom_boxplot()

ggplot(weight_df, aes(x=Gender, y=Height, fill=Gender)) + 
  geom_boxplot()


ggplot(weight_df, aes(x=Weight, fill=Gender)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")

ggplot(weight_df, aes(x=Height, fill=Gender)) +
    geom_histogram(binwidth=.5, alpha=.5, position="identity")


```

### Exercise 1.1e
```{r}
plot(weight_df$Height, weight_df$Weight, col = weight_df$Gender, 
     main = "height over weight; red: men, black: women", xlab = "height", ylab = "weight")
```

### Exercise 1.2b

```{r}
weight_lm_all <- lm(Weight ~ Height, data = weight_df)

# TASK: do the same for weight_df_f and weight_df_m
weight_lm_f <- lm(Weight ~ Height, data = weight_df_f)
weight_lm_m <- lm(Weight ~ Height, data = weight_df_m)
weight_lm_f
weight_lm_m
```

### Exercise 1.2c

```{r}
# TASK
names(weight_lm_f)
names(weight_lm_m)
names(weight_lm_all)
help(lm)
```

### Exercise 1.2d

```{r}
plot(weight_df$Height, weight_df$Weight, col = weight_df$Gender, 
     main = "height over weight; red: men, black: women", xlab = "height", ylab = "weight")
abline(weight_lm_f, col="blue")
abline(weight_lm_m, col="blue")
abline(weight_lm_all, col="green")

ggplot(weight_df, aes(x=Height, y=Weight, color=Gender)) + geom_point(shape=1) +
    scale_colour_hue(l=50) + # Use a slightly darker palette than normal
    geom_smooth(method=lm,   # Add linear regression lines
                se=FALSE,    # Don't add shaded confidence region
                fullrange=TRUE) # Extend regression lines

?abline
weight_lm_all
```

### Exercise 1.2e
Compare the three regression lines in 1.2d and interpret the differences. 

Distributions look very similar. The male distribution is shifted more to the right and top. This is due to the general idea that men are larger and weigh more compared
to women. It is no surprise that the regression lines (male, female) are therefore parrallel to each other. Again the male line is above the female line because of
the general idea that men are larger and weigh more compared to women. Coming to the regression line which refers to all data we can see that it has a higher rate of increase.

### Exercise 1.3a
a) Use `summary(weight_lm_x)` to explore the three linear models:
    * their residuals 
    * the coefficients of the model with standard error, t- and p-values of the statistical hypothesis test
    * residual standard error and R-squared (squared correlation)
    Interpret the p-values in terms of rejecting or not rejecting the null hypothesis of a parameter being zero.  
```{r}
summary(weight_lm_f)
summary(weight_lm_m)
summary(weight_lm_all)
```

Assuming that alpha = 0,05:
- weight_lm_f we reject the null hypothesis. 2e-16
- weight_lm_m we reject the null hypothesis. 2e-16
weight_lm_all we reject the null hypothesis. 2e-16

### Exercise 1.3b
b) Look at the confidence intervals of the linear models using `confint()`. Change the confidence levels from the default 0.95, see how the confidence intervals change and explain why (use the parameter `level = ...`).

```{r}
confint(weight_lm_f)
confint(weight_lm_m)
confint(weight_lm_all)

confint(weight_lm_all, level=0.9)
confint(weight_lm_all, level=0.8)
confint(weight_lm_all, level=0.7)
```