---
title: "Exercise Sheet 2 -- Data Mining <BR> Wirtschaftsinformatik, HTW Berlin"
author: "Martin Spott"
date: "last revision `r format(Sys.Date(), format='%d %B %Y')`"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

The exercises are about multiple linear regression. The data in `miete03.asc` has kindly been provided by the Institute for Statistics at the University of Munich. Import the data set into R using 

```{r}
mietspiegel <- read.table("2uebung_jonas/miete03.asc", header = TRUE)
mietspiegel
```

The data description:

* `nm`:	Net rent in EUR
* `nmqm`:	Net rent per m² in EUR
* `wfl`:	Floor space in m²
* `rooms`:	Number of rooms in household
* `bj`:	Year of construction
* `bez`:	Urban district
* `wohngut`:	Good residential area? (Y=1,N=0)
* `wohnbest`:	Very good residential area? (Y=1,N=0)
* `ww0`:	Hot water supply? (Y=0,N=1)
* `zh0`:	Central heating? (Y=0,N=1)
* `badkach0`:	Tiled bathroom? (Y=0,N=1)
* `badextra`:	Supplementary equipment in bathroom? (Y=1,N=0)
* `kueche`:	Well equipped kitchen? (Y=1,N=0)

The aim of building a linear regression model is to estimate the rent of an accommodation (house/flat) based on data. 

## Exercise 2.1

a) Explore the data set using `summary()` and `View()`. Note that `View()` does not work when you compile (knit) a RMarkdown document, but only in RStudio.
b) Plot the data using `pairs()`, i.e. a matrix of pairwise scatter plots of the variables. Interpret the results.
c) Compute the correlation matrix and show the lower half only for better readability. What do the numbers tell us?
    ```{r}
      miet_cor <- cor(mietspiegel)
      miet_cor[upper.tri(miet_cor, diag = FALSE)] <- NA
      miet_cor
    ```   
    Be aware that `cor()` uses Pearson correlation as a default, which is only sensible for metric features. In particular, it does not make sense for nominal features like `bez` (district). It can be used on binary features like `wohngut` that have an order (Y=1 is better than N=0). We will discuss these issues in the lectures. 
d) Build a multiple linear regression model `miet_lm` with `nm` (*Nettomiete*) as target (output) variable and all the others as predictors (input variables). 
e) Explore the model using `summary()`. Try to make sense of  
    i) the residuals
    ii) the value of the coefficients
    iii) the p-value $Pr(>|t|)$ of the coefficients
    iv) R-squared
    v) the F-statistic and its p-value
f) Why are some of the coefficients in the model negative? Does that make sense? Refer to the data description above for help.


## Exercise 2.2

Think about the following scenario. You would like to buy a flat. You gather data about it through a viewing and information from the estate agent. You want to use the data to estimate the price of the flat. For that purpose you build a multiple linear regression model.  

a) Revisit the predictor (input) variables in your model from Exercise 2.1. Should we use all of them? Do we know all of them in our scenario? 
b) Rebuilt the linear model without the one predictor variable we do not know in our scenario. Compare the model with the one including all predictors. 
c) Interpret the coefficient of the variable `bez` describing the district of Munich. Does this make sense? Check out the data description. 
d) Use the function `factor()` to change the variable `bez`. What does this do? Re-build the model. Compare the model with the previous one and interpret the differences. Look at the values of the coefficients for `bez`. 

## Exercise 2.3

What does the following code fragment do? Explain the value of R-squared and make the link to Exercise 2.2a.
```{r}
mietspiegel_lm <- lm(nm ~ nmqm:wfl, mietspiegel)
summary(mietspiegel_lm)
```



## Solutions

### Exercise 2.1a
```{r}
summary(mietspiegel)
# View(mietspiegel) # this does not work in RMarkdown, only in RStudio
head(mietspiegel)
```

### Exercise 2.1b
```{r}
# plot pairwise scatter plots; sample 500 points only to reduce complexity
pairs(mietspiegel[sample(1:2053, 500),], cex = 0.1)

# Net rent (nm) and floor space (wfl) in qm correlate
# Very good residential area (wohnbest) does have high rent (n), high net rent per m2 (nmqm)
# Good ressidential area (wohngut) does have both sides (high rent, low rent, etc.)
# Most newer apartments lay in a specific range of net rent (nm). Very old buildings have the full range
```
Do we see something interesting?

### Exercise 2.1c
Be aware of the feature types when computing correlation. The function `cor()` uses *Pearson* correlation as default which is suitable for metric features like `nm` and `nmqm`. If we assume that each additional room adds the same to the rent, then *Pearson* correlation can be used to correlate a metric feature like `nm` with `rooms`. However, generally either *Kendall's* or *Spearman's* rank correlations are used for ordinal features like `rooms`. These are available through the parameter `method =` in the function `cor()`. Check the help pages for more information. 

You can see that *Pearson's* and *Spearman's* correlations often give similar results for ordered features like `rooms`. This is also true for binary features like `badextra`. To investigate the relationship between a binary (or any nominal feature) and metric feature, you can also compare boxplots or the confidence intervals of average values.  

```{r}
library(ggplot2)
cor(mietspiegel$nm, mietspiegel$rooms, method = c("pearson")) 
cor(mietspiegel$nm, mietspiegel$rooms, method = c("kendall"))
cor(mietspiegel$nm, mietspiegel$rooms, method = c("spearman"))

cor(mietspiegel$nm, mietspiegel$badextra , method = c("pearson"))
cor(mietspiegel$nm, mietspiegel$badextra, method = c("kendall")) # you want this one 
cor(mietspiegel$nm, mietspiegel$badextra, method = c("spearman")) # or this one

ggplot(mietspiegel, aes(x=badextra, y=nm, group=badextra)) + # or this one
  geom_boxplot()
```


### Exercise 2.1d/e
d) Build a multiple linear regression model `miet_lm` with `nm` (*Nettomiete*) as target (output) variable and all the others as predictors (input variables). 
```{r}
# try all features
# note use of the dot operator `.` representing all remaining variables
mietspiegel_lm <- lm(nm ~ ., data = mietspiegel)
summary(mietspiegel_lm)
```
Explore the meaning of the R output above.
i) the residuals
ii) the value of the coefficients
iii) the p-value $Pr(>|t|)$ of the coefficients
iv) R-squared
v) the F-statistic and its p-value

### Exercise 2.1f
f) Why are some of the coefficients in the model negative? Does that make sense? Refer to the data description above for help.



### Exercise 2.2a
`nmqm` is usually not known. 

### Exercise 2.2b

```{r}
# remove nmqm; note the use of `-nmqm`
mietspiegel_lm2 <- lm(nm ~ .-nmqm, data = mietspiegel)
summary(mietspiegel_lm2)
```
Compare the quality of this model with the one including `nmqm` and explain the differences. 


### Exercise 2.2c
 

### Exercise 2.2d

```{r}
mietspiegel$bez_fact <- factor(mietspiegel$bez)
mietspiegel_lm3 <- lm(nm ~ .-nmqm-bez, data = mietspiegel)
summary(mietspiegel_lm3)
```
Compare the quality of this model with the one using the original column `bez`.

### Exercise 2.3


